Differential invariant signatures for planar Lie group transformations of images

### 2020-07-29 paper/newpaper.tex
edits

@@

@@

\section{Differential invariants through moving frames}

@@

\subsection{$SE(2)$}

\todo[inline]{Make precise!}
Ignoring translation, the action of $SE(2)$ on the first derivatives is simply rotation by some angle $\theta$, i.e.,

\begin{equation}
\left( \begin{array}{c} f_x \\ f_y \end{array} \right) = \left( \begin{array}{cc} \cos \theta & \sin \theta \\ - \sin \theta & \cos \theta \end{array} \right) \left( \begin{array}{c} \bar f_{\bar x} \\ \bar f_{\bar y} \end{array} \right).
\end{equation}

Using $c$ for $\cos \theta$ and $s$ for $\sin \theta$ and prolonging the group to second order via the chain rule gives:

\begin{equation}
\arraycolsep=1.4pt
\left(
\begin{array}{c}
  \bar f_x \\ \bar f_y \\ \bar f_{xx} \\ \bar f_{xy} \\ \bar f_{yy} 
 \end{array}
 \right)
 = 
\left(
\begin{array}{ccccc}
 c & -s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & cs & c^2 - s^2 & -cs \\
0 & 0 & s^2 & 2cs & c^2 \\
\end{array}
\right)\left(
\begin{array}{c}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
 \end{array}
 \right).
\label{eqn:SE2}
 \end{equation}

As $SE(2)$ does not include scaling, lengths stay the same, and so an appropriate choice for first-order normalisation is $\bar f_x = \sqrt(f_x^2 + f_y^2),\:\: \bar f_y = 0$, which sets $c = f_x / \| \nabla f \|$, $s = - f_y / \| \nabla f \|$. This provides the invariant $I_1 = \nabla f$. As there are no more free parameters, the expressions for the second-order terms can be read off as:

\begin{align}
I_2 &= f_x^2 f_{xx} + 2 f_x f_y f_{xy} + f_y^2 f_{yy} \\
I_3 &= f_x f_y (f_{yy} - f_{xx}) + f_{xy} (f_x^2 - f_y^2) \\
I_4 &= f_y^2 f_{xx} - 2 f_x f_y f_{xy} + f_x^2 f_{yy} 
\end{align}

\todo[inline]{ Two things to note: (1) SE(2) has similar invars, (2) there is an issue in E(2) with the moving frame: reflection ambiguity. That's why we did it by contraction.
}

>>>>

h

@@

/subsection{Transformation notation}
Most of the time we will work on grey-scale transformations of two-dimensional images. In this setting, we define the transformation group action by a coordinate transformation $\varphi \colon \mathbb{R}^2 \to \mathbb{R}^2$
\begin{equation}\label{eq:transformation}
  \mathbf{x} \mapsto \bar{\mathbf{x}} = \varphi(\mathbf{x})
\end{equation}
The transformed image is then defined to be
\begin{equation}\label{eq:fbarequalsf}
  \bar{f}(\bar{\mathbf{x}}) = (f\circ\varphi^{-1})(\bar{\mathbf{x}}) = f(\mathbf{x}).
\end{equation}
Derivatives of $\bar{f}$ and $f$ can be related by the chain rule, for example the first order derivatives are given by:
\begin{equation}\label{eq:prolongation}
  \begin{aligned}
    f_x &= \bar{f}_{\bar{x}} \bar{x}_x + \bar{f}_{\bar{y}}\bar{y}_x \\
    f_y &= \bar{f}_{\bar{x}} \bar{x}_y + \bar{f}_{\bar{y}}\bar{y}_y \\
  \end{aligned}
\end{equation}
where $\bar{x}_x = \frac{\partial \varphi_1}{\partial x}(x, y)$, et cetera. It is often convenient to work with $\varphi^{-1}$ instead, which we will do in \S\ref{sec:movingframes}.

@@

\section{Differential invariants through moving frames}\label{sec:movingframes}

@@

\subsection{$SE(2)$}
An element of the transformation group $SE(2)$ acts on an element $(x, y)$ of $\mathbb{R}^2$ to produce $(\bar{x}, \bar{y})$ as
\begin{equation*}
  \begin{aligned}
    \bar{x} &= x\cos\theta  - y\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
Prolonging the group action up to second derivatives via the chain rule computations gives
\begin{equation}
  \begin{bmatrix}
  \bar f_{\bar x} \\ \bar f_{\bar y} \\ \bar f_{\bar{x}\bar{x}} \\ \bar f_{\bar{x}\bar{y}} \\ \bar f_{\bar{y}\bar{y}}
  \end{bmatrix}
  \begin{bmatrix}
 c & -s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & cs & c^2 - s^2 & -cs \\
0 & 0 & s^2 & 2cs & c^2 \\
  \end{bmatrix}
  \begin{bmatrix}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
  \end{bmatrix}
\label{eqn:SE2}
 \end{equation}
where $c = \cos \theta$ and $s = \sin \theta$.

We normalise in two stages. First, we set $\bar{f}_{\bar{y}} = 0$. This determines $\theta$, giving $\cos\theta = f_x/\lVert \nabla f \rVert$ and $\sin\theta = -f_y / \lVert \nabla f \rVert$. We then choose $t_x = -x\cos\theta + y\sin\theta$ and $t_y = -x\sin\theta - y \cos\theta$ to make $\bar{x} = 0$ and $\bar{y} = 0$. These three normalisations fully specify the moving frame, and so the remaining transformed derivatives are invariants. Apart from squaring $I_1$, these can be read off as:
\begin{align}
  I_0 &= f \\
  I_1 &= f_x^2 + f_y^2 \\
  I_2 &= f_x^2 f_{xx} + 2 f_x f_y f_{xy} + f_y^2 f_{yy} \\
  I_3 &= f_x f_y (f_{yy} - f_{xx}) + f_{xy} (f_x^2 - f_y^2) \\
  I_4 &= f_y^2 f_{xx} - 2 f_x f_y f_{xy} + f_x^2 f_{yy} 
\end{align}

\todo[inline]{c.f. transvectant formulae}

\subsection{The Euclidean Group E(2)}
An element of the Euclidean group $E(2)$ maps an element $(x, y)$ of $\mathbb{R}^2$ to $(\bar{x}, \bar{y}$ as 
\begin{equation*}
  \begin{aligned}
    \bar{x} &=  x\varepsilon\cos\theta  - y\varepsilon\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
where $\varepsilon \in \{-1, 1\}$ and the remaining parameters are the same as for $SE(2)$. If $\varepsilon = 1$ it is a rigid transformation, and if $\varepsilon = -1$ it contains a reflection.

This time, the group action is not free on $J^1$ (with elements of the form $(x, y, f, f_x, f_y)$) because there are two possible transformations that map to the normalised coordinates $(0, 0, \bar{f}, \bar{f}_{\bar{x}}, 0)$ where $\bar{f}_{\bar{x}} \ge 0$, one that reflects the gradient, and one that rotates. And first derivatives give no information about $\varepsilon$.

So, we prolong to second derivatives, giving
\begin{equation}
  \begin{bmatrix}
  \bar f_{\bar x} \\ \bar f_{\bar y} \\ \bar f_{\bar{x}\bar{x}} \\ \bar f_{\bar{x}\bar{y}} \\ \bar f_{\bar{y}\bar{y}}
  \end{bmatrix}
  \begin{bmatrix}
 \varepsilon c & -\varepsilon s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & \varepsilon cs & \varepsilon(c^2 - s^2) & -\varepsilon cs \\
0 & 0 & s^2 & 2cs & c^2 \\
  \end{bmatrix}
  \begin{bmatrix}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
  \end{bmatrix}
\label{eqn:E2prolongation}
 \end{equation}
